\section{Thoughts}


\subsection{Predicative Church Encodings}

Booleans:
$$
\begin{array}{llll}
    B_i:
    &\Any_{i+1}
    &:= &
    \Pi A^{\Any_i}. \underbrace{A \to A \to A}_{:\Any_i}
    %
    \\
    \text{true}_i:
    & B_i
    &:=&
    \lambda A^{\Any_i} a^A b^A := a
    %
    \\
    \text{false}_i:
    & B_i
    &:=&
    \lambda A^{\Any_i} a^A b^A := b
    %
    \\
    \text{and}_i:
    & B_{i+1} \to B_i \to B_i
    &:=&
    \lambda a^{B_{i+1}} b^{B_i}. a B_i b \text{ false}_i
\end{array}
$$

Natural numbers:
$$
\begin{array}{llll}
    N_i:
    &\Any_{i+1}
    &:= &
    \Pi A^{\Any_i}. \underbrace{A \to (A \to A) \to A}_{:\Any_i}
    %
    \\
    \text{zero}_i: & N_i &:= &\lambda A^{\Any_i} s^A f^{A \to A}. s
    %
    \\
    \text{succ}_i: & N_i \to N_i &:= &
    \lambda n^{N_i} A^{\Any_i} s^A f^{A \to A}. f (n A s f)
    \\
    %
    \text{plus}_i:
    & N_{i+1} \to N_i \to N_i
    &:=&
    \lambda a^{N_{i+1}} b^{N_i}. a N_i b \text{ succ}_i
    \\
    %
    \text{mult}_i:
    & N_{i+1} \to N_{i+1} \to N_i
    &:=&
    \lambda a^{N_{i+1}} b^{N_{i+1}}
    . a N_{i+1} \text{ zero}_i \underbrace{(\text{plus}_i b)}_{: N_i \to N_i}
\end{array}
$$


Ackermann Function:
$$
a m n :=
\dcases{
a 0 n &:=& 1 + n
\\
a (1+m) 0 &:=& a m 1
\\
a (1+m) (1+n) &:=& a m (a (1+m) n)
}
$$

It can be defined in the impredicative style:
$$
\begin{array}{llll}
    g:
    & (N \to N) \to (N \to N)
    &:=&
    \lambda f n. \text{ succ } n\, N (\text{succ zero}) f
    \\
    %
    a: & N \to N \to N
    &:=&
    \lambda m. m (N \to N) \text{ succ } g
\end{array}
$$









\subsection{STLC Weak Normalization}
%-------------------------------------------------------------------------------


In simply typed lambda calculus all products $\Pi x^A. B$ have the form $A \to
B$ i.e. $x \notin \FV(B)$. Therefore all types can be constructed by the grammar
$$
T ::= X \mid T \to T
$$
where $X$ ranges over (type) variables and $T$ over types. All types $T$ have a
degree $D(T)$ which is defined as
$$
D(T) := \dcases {
    D(X) &:=& 0
    \\
    D(A \to B) &:=& 1 + \text{max}(D(A), D(B))
}
$$

The degree of a redex is the degree of the type of its rator. The degree of a
redex is by definition greater than zero (the rator of a redex must have type $A
\to B$.


Any term $t$ not in normal form has some redexes. For a term we define the
measure
$$
n_{m-1} \ldots n_2 n_1 n_0
$$
where $n_i$ is the number of redexes of degree $i+1$ and $m$ is the greatest
degree if all redexes of the term. The lexicographic order measures is a
wellorder.

\begin{theorem}
    \emph{A rightmost reduction of a term reduces its measure.}

    \begin{proof}
        Let
        $$
        t = C[(\lambda x^A.e) a] \reduce C[e[a/x]]
        $$
        be a rightmost reduction i.e. $(\lambda x^A.e)a)$ is the rightmost redex
        of the term and has degree $k+1$.

        Since the contracted redex is the rightmost redex $e$ and $a$ must be
        in normal form. Therefore all redexes in $e[a/x]$ have to be new redexes
        which have not been present in the original term.

        All newly created redexes have a degree strictly smaller than $k+1$.
        Reason: There are only three possibilities to create new redexes and the
        degree of the new redexes is strictly smaller:
        \begin{enumerate}
            \item The reduction of the rightmost redex has the form
                $$
                (\lambda x^{A \to B}. \ldots x a \ldots ) (\lambda y^A.e)
                \reduce
                \ldots (\lambda y^A. e) a \ldots
                $$
                where $x a$ is in the head position of an application (otherwise
                it were not a new redex). The degree of the newly created redex
                is $D(A \to B)$. The degree of the original redex is $D((A \to
                B) \to C)$ for some type $C$ i.e. the degree of the original
                redex is striclty greater than the degree of the newly created
                redex.

            \item The reduction of the rightmost redex has the form
                $$
                (\lambda x^{A \to B}. x) (\lambda y^A. e)
                \reduce
                \lambda y^A. e
                $$
                and the redex in the original term is applied to an argument
                $a$. I.e. the newly created redex is $(\lambda y^A. e) a$ and
                has degree $D(A \to B)$. The original redex has degree $D((A \to
                B) \to (A \to B))$ which is strictly greater.


            \item The reduction of the rightmost redex has the form
                $$
                (\lambda x^A. (\lambda y^B. e)) a
                \reduce
                \lambda y^B. e[a/x]
                $$
                and the redex in the original term is applied to an argument
                $b$. I.e. the newly created redex is $(\lambda y^B. e[a/x]) b$
                and has degree $D(B \to C)$. The original redex has degree $D(A
                 \to (B \to C))$ which is strictly greater.
        \end{enumerate}

        Therefore the number of degree $k$ redexes decreases by one, the numbers
        of redexes with higher degree stay the same. I.e. the measure decreases.
    \end{proof}
\end{theorem}

\begin{remark}
    In the proof of the previous theorem the requirement that the redex is
    rightmost can be relaxed a little bit. It is at least necessary that
    the argument $a$ of the redex $(\lambda x^A.e) a$ only contains
    redexes of lower degree. Since the bound variable $x$ might be in the
    body more than once, the contractum $e[a/x]$ might multiply the number of
    redexes of $a$. As long as they are of lower degree, the measure decreases.
\end{remark}







\subsection{Confluence with Marks}
%-------------------------------------------------------------------------------
{

    \begin{definition}
        \emph{Terms with marked redexes.}
        $$
        t ::= x \mid t t \mid \lambda x. t \mid (\lambda' x. t) t
        $$
    \end{definition}



    \begin{definition}
        \emph{Remove all marks.}
        The term
        $$
        |t|
        $$
        is the term $t$ where all marked $\lambda'$ are replaced by $\lambda$.
    \end{definition}




    \begin{lemma}
        \label{thm:ReductionLiftedToMark}
        $$
        \ruleh{
            |a'| = a
            \quad
            a \reduce b
        }
        {
            \existbracket {b'}{
                a' \reduce b'
                \\
                |b'| = b
            }
        }
        $$

        \begin{proof}
            Easy. Just contract the corresponding redex in $a'$ to get $b'$.
        \end{proof}
    \end{lemma}




    \begin{definition}
        \emph{Reduce all marked redexes.}
        $$
        \varphi(t) :=
        \dcases{
            \varphi(x) &:=& x
            \\
            \varphi(a b) &:=& \varphi(a) \varphi(b)
            \\
            \varphi(\lambda x .e) &:=& \lambda x. \varphi(e)
            \\
            \varphi ((\lambda x.e) a) &:=& \varphi(e)[\varphi(a)/x]
        }
        $$
    \end{definition}




    \begin{lemma}
        \label{thm:PhiSubstitution}
        $$
        \varphi(e[a/x]) = \varphi(e)[\varphi(a)/x]
        $$

        \begin{proof}
            By induction on the structure of $e$
        \end{proof}
    \end{lemma}




    \begin{lemma}
        \label{thm:PhiReduction}
        $$
        \ruleh{
            a \reduce b
        }
        {
            \varphi(a) \reduce \varphi(b)
        }
        $$

        \begin{proof}

            By induction on $a \reduce b$?
            % Check, if one step reduction is ok in the conclusion!!
            %------------------------------------------------------------

            MISSING!!
        \end{proof}
    \end{lemma}



    \begin{lemma}
        \label{thm:ErasePhi}
        $$
        |a| \reducestar \varphi(a)
        $$

        \begin{proof}
            By induction on the structure of $a$.

            MISSING!!
        \end{proof}
    \end{lemma}






    \begin{lemma}
        \label{thm:StripLemma}
        \emph{Strip lemma.}
        %------------------------------------------------------------
        $$
        \begin{array}{lll}
            a & \reducestar & b
            \\
            \downarrow_\beta & & \downarrow_{\beta^*}
            \\
            c & \reducestar & \exists d
        \end{array}
        $$

        \begin{proof}
            Let $(\lambda x.e)f$ be the redex in $a$ which is contracted in $a
            \reduce c$ and $a'$ the term $a$ where the contracted redex is
            replaced by $(\lambda' x. e)f$.

            Since $a'$ has only one marked redex we get
            $$
            c = \varphi(a')
            $$

            Repeated application of~\ref{thm:ReductionLiftedToMark} on $a
            \reducestar b$ guarantees the existence of a term $b'$ with
            $$
            \vertlist{
                |b'| = b
                \\
                a' \reducestar b'
            }
            $$

            Repeated application of~\ref{thm:PhiReduction} on $a' \reducestar
            b'$ gives us
            $$
            c = \varphi(a') \reducestar \varphi(b')
            $$

            Moreover from \ref{thm:ErasePhi} we have
            $$
            b = |b'| \reducestar \varphi(b')
            $$
            i.e. $d := \varphi(b')$ satisfies the requirements of the theorem.
            $$
            \begin{array}{ccc}
                a = |a'| & \reducestar & b = |b'|
                \\
                \downarrow_\beta & & \downarrow_{\beta^*}
                \\
                c = \varphi(a') & \reducestar &  d = \varphi(b')
            \end{array}
            $$
        \end{proof}
    \end{lemma}




    \begin{theorem}
        \label{thm:Confluence}
        \emph{Confluence.}
        $$
        \begin{array}{lll}
            a & \reducestar & b
            \\
            \downarrow_{\beta^*} & & \downarrow_{\beta^*}
            \\
            c & \reducestar & \exists d
        \end{array}
        $$

        \begin{proof}
            By repeated application of the strip lemma~\ref{thm:StripLemma} on
            $a \reducestar c$.
        \end{proof}
    \end{theorem}



}







\subsection{Barendregt-Geuvers-Klop Conjecture}
%-------------------------------------------------------------------------------

Proof goals:
\begin{enumerate}
    \item If a welltyped term has a normal form, then it can be found
        by a rightmost reduction.

    \item If a rightmost reduction of a term ends in a normal form, then
        the term is strongly normalizing.

    \item  If all subterms of a term are weakly normalizing, then the term
        is strongly normalizing. Or the other way round: If a term is not
        strongly normalizing, then it must have a subterm which is not
        normalizing.

        It might be necessary to make the premise stronger: If all subterms of
        any reduct of a term are weakly normalizing, then the term is strongly
        normalizing.

        This formulation might be better fitted to typed lambda calculi
        where we try to prove that all welltyped terms are weakly normalizing.
        Due to subject reduction all reducts of a welltyped term are welltyped
        as well.
\end{enumerate}

Candidates for intermediate lemmas:
\begin{enumerate}
    \item A rightmost reduction reduces all redexes of a term.

        Proof idea: A rightmost reduction step contracts the rightmost redex.
        The contraction only modifies subterms of all redexes
        in which is is contained. But the redexes remain. Finally it has to
        contract all of them.
\end{enumerate}


Concept: Minimal terms with an infinite reduction sequence where all proper
subterms are strongly normalizing. Examples $(\lambda x. x x)(\lambda x. x x)$,
$U U$ where $U = \lambda x y. y (x x y)$. Counterexamples: $\lambda x. UU$ is
not minimal, $UU f$ has a proper subterm $UU$ which is not strongly normalizing.

Such terms must be applications with an abstraction in the head position.
Reason: All subterms can be converted to their normal form.  If the term was not
an application with  an abstraction in the head position, it would be normal and
could not have an infinite reduction sequence.

Note: A rightmost reduction reduces all subterms of a redex before it contracts
a redex. I.e. a rightmost reduction will finally encounter all terms with an
infinite reduction sequence which have only strongly normalizing proper
subterms.

All terms with an infinite reduction sequence (i.e. not strongly normalizing)
have minimal substerms with an infinite reduction sequence where all proper
subterms have no infinite reduction sequence i.e. where all proper subterms are
strongly normalizing. I.e. we can concentrate on terms with an infinite
reduction sequence where all proper subterms are strongly normalizing. If such
welltyped terms do not exist, then all terms are strongly normalizing.


\begin{theorem} If a term $t$ has an infinite reduction sequence, then the
    rightmost reduction sequence is infinite.

    \begin{proof}
        Main idea: A rightmost reduction of a term with an infinite reduction
        sequence never ends. After a finite number of steps it reduces to a term
        where the rightmost reduction continues.

        $t$ has  a rightmost subterm $r$ which has an infinite reduction
        sequence. Thus all proper subterms of $r$ and all terms to the right of
        $r$ are strongly normalizing. The rightmost reduction sequence reduces
        all terms to the right of $r$ and all proper subterms of $r$ to their
        normal form.

        $r$ must be an application of the form $(\lambda x. e) a \vec b$. This
        is the only possible form where all proper subterms are normal and the
        term still reduces.

        The rightmost reduction reduces this term to $e[a/x] \vec b$. $e$, $a$
        and $\vec b$ are in normal form. Therefore the term $e[a/x] \vec b$ must
        contain newly created redexes.
    \end{proof}
\end{theorem}







\begin{definition}
    \emph{Call by value reduction} $a \betarv b$ defined by the compatible
    closure of the rule
    $$
    \ruleh{
        a \in \NF
    }
    {
        (\lambda x^A. e) a \betarv e[a/x]
    }
    $$
\end{definition}






\begin{definition}
    \emph{$n$-step relation}
    \begin{enumerate}
        \item $$ a \toreln r 0 a$$

        \item $$
            \ruleh{
                a \torel r b
                \\
                b \toreln r i c
            }
            {
                a \toreln r {i+1} c
            }
            $$
    \end{enumerate}
\end{definition}






\begin{definition}
    \emph{$n$-step beta reduction}
    \begin{enumerate}
        \item $$ a \betarn 0 a$$

        \item $$
            \ruleh{
                a \reduce b
                \\
                b \betarn i c
            }
            {
                a \betarn {i+1} c
            }
            $$
    \end{enumerate}
\end{definition}




\begin{definition}
    \emph{Evaluation context.} An evaluation context is a lambda term with one
    hole. It is defined by the grammar
    $$
    E ::= [.] \mid t E \mid E t \mid \lambda x. E
    $$
    Where $t$ ranges over lambda terms and $E$ ranges over evaluation contexts.
\end{definition}




\begin{definition}
    \emph{Perpetual redex.}
    A redex $r = \redex x A e a$ is perpetual if its contraction within any
    diverging term $C[r]$ implies that $C[e[a/x]]$ is diverging.
\end{definition}




\begin{theorem}
    \label{thm:NonLeftmostReduction}
    \emph{Non leftmost reduction step.}
    %--------------------------------------------------
    Let $t = C[(\lambda x^A. e)a]$ be a term where $(\lambda x^A.e)a$ is the
    leftmost redex and $t$ reduces to $u$ in one step without contracting its
    leftmost redex. Then there exist some reducts $A'$, $e'$, $a'$ and $C'$ of
    $A$, $e$, $a$ and $C$ such that $u = C'[(\lambda x^{A'}. e') a']$.
    $$
    \rulev{
        t \torel \betanl u
        \\
        t = C[(\lambda x^A. e) a]
        \\
        (\lambda x^A. e) a \text{ is leftmost}
    }
    {
        \existbracket {C' e' a'}
        {
            A \reducestar A'
            \\
            e \reducestar e'
            \\
            a \reducestar a'
            \\
            C \reducestar C'
            \\
            u = C'[(\lambda x^{A'}. e') a']
            \\
            (\lambda x^{A'}. e') a' \text{ is leftmost}
        }
    }
    $$
    \begin{proof} The fact is quite evident. Since the leftmost redex has not
        been contracted some other redex in one of the subterms must have been
        contracted. The other redex must be either in one of the subterms of the
        leftmost redex or in another subterm outside the redex.
    \end{proof}
\end{theorem}




\begin{theorem}
    %  lambda I redex
    %------------------------------------------------------------
    $$
    \rulev{
        C[(\lambda x^A. e) a)] \text{ is diverging}
        \\
        (\lambda x^A. e) a \text{ is leftmost}
        \\
        x \notin \FV(e)
        \\
        A, a \in \SN
    }
    {
        C[e] \text{ is diverging}
    }
    $$
    \begin{proof}
        Let
        $$
        t_0 \reduce t_1 \reduce t_2 \ldots
        $$
        be an infinite reduction sequence starting at $t_0 = C[(\lambda x^A.e)
        a]$. Such a reduction sequence exists because $t_0$ is diverging. We
        distinguish two cases
        \begin{enumerate}
            \item All reduction steps are not leftmost:

                Then by the theorem
                about non leftmost reduction~\ref{thm:NonLeftmostReduction} we
                have $t_i = C_i[(\lambda x^{A_i}. e_i) a_i]$ where the redex
                remains the leftmost redex and in each reduction step either a
                proper subterm of the redex is reduced or a term to the
                right of the redex is reduced.

                Since $A$ and $a$ are strongly normalizing there has to be a $k$
                such that $A_k = K_j $ and $a_k = a_j$ for all $j \ge k$.
                Therefore from $k$ on in each reduction step either $e$ is
                reduced or some other term to the left of the redex is reduced.

                All the subterms reduced from step $k$ on are present in
                $C_j[(\lambda x^{A_j}. e_j) a_j]$ and $C_j[e_j]$ for $k \le j$.
                Therefore
                $$
                C_k[e_k] \reduce C_{k+1}[e_{k+1}] \reduce \ldots
                $$
                is an infinite reduction sequence and since $C \reducestar
                C_k$ and $e \reducestar e_k$ there is the infinite reduction
                sequence
                $$
                C[e] \reducestar
                C_k[e_k] \reduce C_{k+1}[e_{k+1}] \reduce \ldots
                $$

            \item There is a first leftmost reduction step $t_k \torel
                \betal t_{k+1}$ for some $k$:

                Then the infinite reduction sequence
                $$
                t_0
                \torelstar \betanl
                t_k
                \torel \betal
                t_{k+1}
                \reduce
                t_{k+2}
                \reduce
                \ldots
                $$
                %
                by~\ref{thm:NonLeftmostReduction}
                must have the form
                %
                $$
                C[(\lambda x^{A}. e) a]
                \torelstar \betanl
                C_k[(\lambda x^{A_k}. e_k) a_k]
                \torel \betal
                C_k[e_k]
                \reduce
                t_{k+2}
                \reduce
                \ldots
                $$

                Since $C \reducestar C_k$ and $e \reducestar e_k$ we have $C[e]
                \reducestar C_k[e_k]$ and therefore
                $$
                C[e] \reducestar C_k[e_k] \reduce t_{k+2} \reduce \ldots
                $$
                is an infinite reduction sequence.
        \end{enumerate}
    \end{proof}
\end{theorem}

{
    \def\redex#1#2#3#4{{(\lambda #1^{#2}. #3) #4}}




    \begin{theorem}
        \def\red{{\redex x A e a}}
        \def\redi#1{{\redex {x_{#1}} {A_{#1}} {e_{#1}} {a_{#1}}}}
        \def\contr{{e[a/x]}}

        $$
        \rulev{
            C[\red] \text{ is diverging}
            \\
            \red \text{ is leftmost}
            \\
            x \in \FV(e)
            \\
            A \in \NF
        }
        {
            C[\contr] \text{ is diverging}
        }
        $$
        \begin{proof}
            Let
            $$
            C[\red] = t_0 \reduce t_1 \reduce \ldots
            $$
            be an infinite reduction sequence which exists because $t_0$ is
            diverging. We distinguish two cases:
            \begin{enumerate}

            \item All reductions are not leftmost. Therefore the redex is never
                contracted in the infinite reduction sequence.

                By the theorem~\ref{thm:NonLeftmostReduction} the reduction
                sequence must have the form
                $$
                    C_0[\redi 0]
                    \reduce C_1[\redi 1]
                    \reduce C_2[\redi 2]
                    \reduce \ldots
                $$
                where in each step one of $C_i$, $e_i$ or $a_i$
                reduces. Since the reduction sequence is infinite at least one
                of them must have an infinite reduction sequence.

                If $a$ is divergent then $C[\contr]$ is divergent because $a$
                    occurs in $C[\contr]$.

                If $e$ is divergent, then $e[a/x]$ is divergent as well because
                    substitution respects reduction ($e \reduce e' \imp e[a/x]
                    \reduce e'[a/x]$).

                The same applies if $C$ is divergent.

            \item There is a first leftmost reduction step $t_k \torel\betal
                t_{k+1}$ for some $k$: In that case the infinite reduction
                sequence looks like
                $$
                    C_0[\redi 0]
                    \torelstar\betanl C_k[\redi k]
                    \reduce C_k[e_k[a_k/x]]
                    \reduce t_{k+2}
                    \reduce \ldots
                $$
                Therefore we have for $C[\contr]$ the infinite reduction
                sequence
                $$
                C[\contr]
                \reducestar C_k[e_k[a_k/x]]
                \reduce t_{k+2}
                \reduce \ldots
                $$
            \end{enumerate}
        \end{proof}
    \end{theorem}
}


\begin{theorem}
    \label{thm:HeadRedexStep}
    $$
    \rulev{
        (\lambda x^A. e) a \vec b \in \SN
        \\
        c \in SN
    }
    {
        (\lambda x^A. e) a \vec b c \in \SN
    }
    $$
    \begin{proof}
        MISSING
    \end{proof}
\end{theorem}





\begin{theorem}
    \label{thm:StronglyNormalizingRedex}
    $$
    \rulev{
        A, e, a, e[a/x] \vec b  \in \SN
    }
    {
        (\lambda x^A. e) a \vec b \in \SN
    }
    $$
    \begin{proof} By induction on the length of $\vec b$.

        The base case $|\vec b| = 0$ is trivial.

        The inductive case is proved by the previous
        theorem~\ref{thm:HeadRedexStep}.

        MISSING: Induction step needs rework!!
    \end{proof}
\end{theorem}








\begin{definition}
    \emph{Diverging term.} A term $t$ is diverging if it has an infinite
    reduction sequence
    $$
    t = t_0 \reduce t_1 \reduce t_2 \reduce \ldots
    $$
\end{definition}





\begin{definition}
    \emph{Minimal diverging term.} A term $t$ is minimal diverging if it is
    diverging and all its proper subterms are strongly normalizing.
\end{definition}




\begin{theorem}
    \label{thm:FormMinDiverging}
    \emph{A minimal diverging term $t$ has the form $(\lambda x^A.e) a \vec b$.}

    \begin{proof}
        This is evident by the following facts:
        \begin{enumerate}
        \item $t$ cannot be a sort or a variable because they are not diverging.

        \item $t$ cannot be a product or an abstraction because they are
            strongly normalizing if their proper subterms are strongly
            normalizing.

        \item Therefore it has to be an application. An application is strongly
            normalizing if its proper subterms are strongly normalizing unless
            its head term is an abstraction.
        \end{enumerate}
    \end{proof}
\end{theorem}

Note that it is perfectly possible that $\vec b$ is not empty in a minimal
diverging term. E.g.
$$
\begin{array}{lll}
    (\lambda x y z. x z) \omega a \omega
    &\reduce&
    (\lambda y z. \omega z) a \omega
    \\
    &\reduce&
    (\lambda z. \omega z) \omega
    \\
    &\reduce&
    \omega \omega
\end{array}
$$
where $\omega = \lambda x. x x$. Without the last argument the term would be
strongly normalizing with its unique normal form $\lambda z. \omega z$.






\begin{theorem}
    \label{thm:MinDivergingLeftMostDiverging}
    %------------------------------------------------------------
    $$
    \rulev{
        \redex x A e a \vec b \text{ is minimal diverging}
    }
    {
        e[a/x] \vec b \text{ is diverging}
    }
    $$
    \begin{proof}
        MISSING
    \end{proof}
\end{theorem}





\begin{theorem}
    WRONG THEOREM!!!

    \emph{All one step reducts $u$ of a minimal diverging term $t$ are
    diverging.}
    $$
    \rulev{
        t \text{ is minimal diverging}
        \\
        t \reduce u
    }
    {
        u \text{ is diverging}
    }
    $$


    \begin{proof}
        This statement is wrong. We construct a counter example:
        $$
        (\lambda x y. x y) (\lambda x. (\lambda y. I) (x x)) \omega
        $$
        This term is diverging as can be shown by the following leftmost
        reduction path:
        $$
        \begin{array}{lll}
            (\lambda x y. x y) (\lambda x. (\lambda z. I) (x x)) \omega
            &\torel{\betal^2}&
            (\lambda x. (\lambda z. I) (x x)) \omega
            \\
            &\torel\betal&
            (\lambda y. I) \Omega \quad \in \DIV
        \end{array}
        $$
        The term is certainly minimal, because without the second argument
        $\omega$ the perpetual strategy ends up in the normal form term $\lambda
        y. I$.
        $$
        \begin{array}{lll}
            (\lambda x y. x y) (\lambda x. (\lambda z. I) (x x))
            &\torel{\betae}&
            \lambda y. (\lambda x. (\lambda z. I) (x x)) y
            \\
            &\torel{\betae}& \lambda y. (\lambda z. I) (y y)
            \\
            &\torel{\betae}& \lambda y. I
        \end{array}
        $$

        However the term reduces in one step to a term which is no longer
        diverging
        $$
        \begin{array}{lll}
            (\lambda x y. x y) (\lambda x. (\lambda z. I) (x x)) \omega
            &\torel\betar&
            (\lambda x y. x y) (\lambda x. I) \omega
            \\
            & & \text{\scriptsize now use perpetual strategy}
            \\
            &\torel{\betae^2}&
            (\lambda x. I) \omega
            \\
            &\torel{\betae}&
            I
        \end{array}
        $$
    \end{proof}
\end{theorem}
